---
title: "Assessment02: Part I: Processing Fuel Prices Data"
subtitle: "DSB100 - Fundamentals of Data Science"
author: "Student name and id: replace with your name and student id, eg Adam Smith, n12345678 "
date: "October 2026"
output:
  html_document:
    toc: true
    toc_float:
      toc_collapsed: true
    toc_depth: 3
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
options(width = 90)
knitr::opts_chunk$set(echo = TRUE, comment = NA, eval = TRUE)
knitr::opts_chunk$set(fig.width = 4.8, fig.height = 3.6)
options(dplyr.summarise.inform = FALSE, tidyverse.quiet = TRUE, width = 120)
```

# About this assessment

This assessment is about working with and making sense of data using R
and fundamental concepts in probability, statistics and visualisation.

It requires you:

- to write and execute code to explore and manipulate data
- to explain things about data, working with data, and data ethics
- to figure things out, including by consulting appropriate sources.

Please check Canvas for more information about this assessment,
including:

- [Instructions](https://canvas.qut.edu.au/courses/23983/pages/assessment-2-instructions "Instructions")
- [When this Assessment is
  due](https://canvas.qut.edu.au/courses/23983/pages/assessment-2-instructions "Due date")
- [Use of AI and other resources in this
  Assessment](https://canvas.qut.edu.au/courses/23983/pages/assessment-policy "AI Policy")
- [How will we run and mark your
  submission](https://canvas.qut.edu.au/courses/23983/pages/assessment-2-marking "Assessment 2: Marking")
- [Errors and corrections to this
  Assessment](https://canvas.qut.edu.au/courses/23983/pages/assessment-2-errata "Assessment 2: Errata").

------------------------------------------------------------------------

# Part I: Processing fuel prices data (30 marks total)

In this part of the assessment, you will be processing fuel prices data
across Queensland. The data file, `fuel_prices_july.csv`, contains the
average fuel prices recorded across a number of petrol stations across
Queensland.

The dataset contains the following columns:

- `site_id`: a unique ID code identifying the site
- `site_brand`: the brand of the petrol station
- `site_suburb`: suburb name for the location of the site
- `site_state`: state name for the location of the site
- `fuel_type`: the type of the fuel recorded for the entry
- `price`: the price of fuel, given in units of 1/10 cent. e.g., 2321
  means \$2.321. An entry of 9999 is used to indicate a missing value.

**Note:** The dataset is adapted from the fuel prices recorded by the
Queensland government treasury. The entries have been pre-processed from
its raw format, and randomised to remove identification. Brand names
have also been changed to fictional ones. More details about the
original dataset can be found
[here](https://www.data.qld.gov.au/dataset/fuel-price-reporting-2026),
which is published under a [CC BY
4.0](https://creativecommons.org/licenses/by/4.0/) license. But do not
use the manual for the original dataset in this assessment since several
conventions have been changed from the original dataset.

## Task and instructions

Your task is to do some initial processing and visualisation of the data
and conduct hypothesis tests to statistically answer some questions of
interest around the differences and associations in fuel prices.

This is broken down into the sequence of tasks listed below, which
largely follows a typical workflow. These questions involve a
combination of writing code and word responses. For each question,
please write your answer under the `Answer` heading provided. Any
required code is to be written within the R code blocks provided, and
any required word responses are to be written in the plain text blocks
provided.

Please do not add additional response blocks for either code or text.

## Clean start

```{r clean-start, class.source="bg-success"}
rm(list = ls())
```

## Load only the specified libraries

```{r load-libraries, warning=FALSE, message=FALSE, class.source="bg-success"}
library(ggplot2)
library(tidyr)
library(dplyr)
library(readr)
```

## Load the dataset

```{r load-data, class.source="bg-success"}
fuel_prices_raw <- read_csv("data/fuel_prices_july.csv")
```

------------------------------------------------------------------------

# Question I.1 (6 marks): Processing the data

Before we start any analysis, let's take a look at the data.

## Question I.1.1 (1 mark)

Looking at the `fuel_prices_raw` tibble, answer the following:

- How many fuel price entries were recorded across the entire dataset?
- What variable type is each column in the tibble stored as?

### Answer I.1.1

```{r QI.1.1-code}
# Write your code here
nrow(fuel_prices_raw)

for (col in colnames(fuel_prices_raw)) {
  print(sprintf("%s data type: %s", col, typeof(fuel_prices_raw[[col]])))
}
```

```{asis QI.1.1-text}
there are 2346 fuel price entries across the entire dataset. 

the data type for each column is stored as the following: 
[1] "site_id data type: double"
[1] "site_brand data type: character"
[1] "site_suburb data type: character"
[1] "site_state data type: character"
[1] "fuel_type data type: character"
[1] "price data type: double"

```

## Question I.1.2 (1 mark)

Make a plot with each fuel price entry in `fuel_prices_raw` as a point
on the plot. Use the `price` variable for the `x` axis and the
`fuel_type` variable as the `y` axis. Remember to use jitter and
transparency to remedy the overplotting.

### Answer I.1.2

```{r QI.1.2-code}
ggplot(data=fuel_prices_raw, aes(x=price, y=fuel_type)) +
  geom_jitter(alpha=0.5) +
  labs(
    title = "Price of Each Fuel Type",
    x = "Price ($)",
    y = "Fuel Type"
  )

```

## Question I.1.3 (1 mark)

Based on your plot from above, comment on the following:

- Are there any prices recorded as 9999? (Remember that these are used
  to indicate a missing entry in the dataset.)
- Are there any other outliers?

### Answer I.1.3

```{asis QI.1.3-text}
There appears to be approx a dozen plots along the $10,000 range across various fuel types, indicating that they are missing entries. Additionally, there appears to be serveral outliers two between $2500 and $5000 seperate from the main clusters.

```

## Question I.1.4 (2 marks)

We will convert any invalid entries into `NA` to properly indicate that
they are indeed invalid. In addition, prices that are too low (e.g.,
less than \$0.5) are likely incorrect entries.

Process the existing `fuel_prices_raw` tibble to:

- Replace any entries of `price` that are equal to `9999` with `NA`.
- Replace any entries of `price` that are less than `500` with `NA`.

Assign this to a new tibble `fuel_prices`.

**Important:** We will work with the processed data in `fuel_prices`
instead of the raw dataset `fuel_prices_raw` from now on, so check your
pre-processing by running the following two lines of code:

- `sum(is.na(fuel_prices$price))`
- `mean(fuel_prices$price, na.rm=TRUE)`

and compare your summaries against these:

```         
> sum(is.na(fuel_prices$price))
[1] 15

> mean(fuel_prices$price, na.rm=TRUE)
[1] 1938.349
```

Provide a quick explanation of what each line of the provided summary
code does.

### Answer I.1.4

```{r QI.1.4-code}
# Write your code here
fuel_prices <- fuel_prices_raw %>%
  mutate(
    price = if_else (price >= 500 & price < 9999, price, NA_real_)
  )
# Uncomment this to confirm your summaries
sum(is.na(fuel_prices$price))
mean(fuel_prices$price, na.rm=TRUE)
```

```{asis QI.1.4-text}
*sum(is.na(fuel_prices$price))*: returns the count of NA in column price of fuel_prices - sum check to see number of NA values

*mean(fuel_prices$price, na.rm=TRUE)*: returns the mean of price column from fuel_prices with optional argument to ignore missing values. 

```

## Question I.1.5 (1 mark)

Make a new tibble of your `fuel_prices` dataset (the cleaned up one) by
converting it into wide format where each `fuel_type` is in its own
column. Assign this to a tibble called `fuel_type_wide`.

That is, the tibble should look something like

| `site_id` | `site_brand` | `site_suburb` | `Diesel` | `PULP 98 RON` | ... |
|:---------:|:------------:|:-------------:|:--------:|:--------------|:---:|
|  XXXXXXX  |  XXXXXXXXXX  |  XXXXXXXXXXX  |  XXXXXX  | XXXXXXXXXXX   | ... |

Retain all the `NA` values from the `fuel_prices` dataset.

Hint: look into `pivot_wider`.

### Answer I.1.5

```{r QI.1.5-code}
# Write your code here
fuel_type_wide <- fuel_prices %>%
  pivot_wider(
    names_from = fuel_type, 
    values_from = price
    
  )

```

------------------------------------------------------------------------

# Question I.2 (10 marks): Exploring the data

Now that you have cleaned up the data, and have them configured in both
the longer and wider formats (`fuel_prices` and `fuel_prices_wider`), we
now proceed to explore the datasets.

## Question I.2.1 (2 marks)

Write some code to answer the following questions:

- Which petrol station in Queensland had the highest Unleaded price?
- Which petrol station in Queensland had the lowest Diesel price?

State both answers in terms of their site ID and suburb.

### Answer I.2.1

```{r QI.2.1-code}

max_value = max(fuel_prices$price, na.rm=TRUE) 
sprintf("Max price: %i", max_value)

max_value = max(fuel_prices$price, na.rm=TRUE) 
sprintf("Max price: %i", max_value)

```

```{asis QI.2.1-text}
Write your response here in plain text

```

## Question I.2.2 (2 marks)

Generate the following summary statistics for Unleaded fuel for each
`site_brand`:

- Mean
- Standard deviation
- First quartile (Q1)
- Median
- Third quartile (Q3)

Briefly compare the brands `Velora` and `AG` in terms of their central
tendency and spread in two or three sentences.

### Answer I.2.2

```{r Q2.2}
# Write your code here 

```

```{asis QI.2.2-text}
Write your response here in plain text

```

## Question I.2.3 (2 marks)

To better understand the differences, create histograms of `price` for
Unleaded fuel for each brand. Facet the histograms by `site_brand` in a
way that they have the same scale for the x-axis for easier comparison.
Use an informative name for the x-axis. Remember to try a few different
bins/binwidth settings and find an appropriate one for this dataset

### Answer I.2.3

```{r QI.2.3-code}
# Write your code here

```

## Question I.2.4 (1 mark)

Make violin plots of `price` on the x-axis separated by `fuel_type` in
the y-axis and overlay semi-transparent jittered points. Note: you might
want to experiment with some different heights and transparency settings
of the jitter as to not obscure the violin plot.

### Answer I.2.4

```{r QI.2.4-code}
# Write your code here 

```

## Question I.2.5 (1 mark)

How many entries recorded Unleaded fuel prices of 200 cents or more?
What is this number as a proportion of the total number of Unleaded fuel
price entries?

### Answer I.2.5

```{r QI.2.5-code}
# Write your code here

```

```{asis QI.2.5-text}
Write your response here in plain text
```

## Question I.2.6 (2 marks)

Using the wider format data, `fuel_prices_wider`, plot the Unleaded
versus Diesel data, with Unleaded prices on the x-axis and Diesel prices
on the y-axis. Use informative axis labels.

Briefly comment on the correlation between the prices in one or two
sentences.

### Answer I.2.6

```{r QI.2.6-code}
# Write your code here 

```

```{asis QI.2.6-text}
Write your response here in plain text
```

------------------------------------------------------------------------

# Question I.3 (7 marks): Comparing prices between brands

We want to know if the fuel brand `Velora` tends to have lower Unleaded
fuel prices compared to another brand `AG`. This does seem to be the
case based on the exploratory analyses above. To make this comparison
more precise, we will conduct a one-tailed t-test to see if the average
Unleaded prices of Velora are indeed lower than AG.

## Question I.3.1 (2 marks)

Briefly explain the following:

- What does the t-test aim to compare?
- Why is a one-tailed test needed?
- Would our test be a one-sample or two-sample test?
- Why is Welch's t-test useful for this dataset?

### Answer I.3.1

```{asis QI.3.1-text}
Write your response here in plain text

```

## Question I.3.2 (1 mark)

Create two numeric vectors: Unleaded prices for Velora and Unleaded
prices for AG.

### Answer I.3.2

```{r Question I.3.2}
# Write your code here

```

## Question I.3.3 (2 marks)

Using the t-test:

- Write H0 and H1.
- State the exact test that you're doing (refer to your response in Q
  I.3.1)
- Run the test in R.

Make sure the test you're doing matches your R code.

### Answer I.3.3

```{asis QI.3.3-text}
Write your response here in plain text

```

```{r QI.3.3-code}
# Write your code here

```

## Question I.3.4 (1 mark)

Interpret the result from the test. What does the p-value imply about
statistical significance in this fuel pricing context? Use one to three
sentences.

### Answer I.3.4

```{asis QI.3.4-text}
Write your response here in plain text

```

## Question I.3.5 (1 mark)

By using the data in the manner above, we have treated each price entry
as an independent observation. This is one of the main assumptions of
the t-test.

What is a reason why the fuel price entries are not actually independent
observations? What is a potential consequence of this?

### Answer I.3.5

```{asis QI.3.5-text}
Write your response here in plain text

```

------------------------------------------------------------------------

# Question I.4 (6 marks): Linear regression modelling

We also want to know if there is a relationship between Unleaded and
Diesel prices. To make this more precise, we will look to fit a linear
regression model.

## Question I.4.1 (1 mark)

Fit a linear regression model with:

- Response variable: Diesel price
- Predictor: Unleaded price

Show a `summary()` for the model.

### Answer I.4.1

```{r QI.4.1-code}
# Write your code here

```

## Question I.4.2 (1 mark)

Explain in a few sentences what the value of the coefficient of
`Unleaded` in the model means and how its $p$ value supports or
contradicts your observation made in Question 2.6.

### Answer I.4.2

```{asis QI.4.2-text}
Write your response here in plain text

```

## Question I.4.3 (3 mark)

- Provide a QQ-plot of the model residuals along with the a line
  corresponding to the diagonal (QQ-line)
- Describe what the plot shows in relation to the assumptions that
  underpin linear regression models.
- Discuss whether any transformation of the response variable is
  warranted.

### Answer I.4.3

```{r QI.4.3-code}
# Write your code here

```

```{asis QI.4.3-text}
Write your response here in plain text
```

## Question I.4.4 (1 mark)

Does your result support the claim that "Unleaded prices *impact* Diesel
prices in Queensland"? Why/why not?

### Answer I.4.4

```{asis QI.4.4-text}
Write your response here in plain text.

```

------------------------------------------------------------------------

# Question I.5 (1 mark): Working reproducibly

Follow the reproducible practices covered in the unit (see the Unit
Canvas page on this topic and the recommended textbook).

Your submission must:

- have a separate file for each part (this is part I, so it should be
  named `Assessment02_Part1.Rmd`)
- render to HTML without error by the Teaching Team;
- use clear code, informative comments, and readable plots;
- avoid hidden dependencies and hard-coded local paths.

------------------------------------------------------------------------

# End of Part I
