---
title: "Assessment02 Part II: Modelling sports performance"
subtitle: "DSB100 - Fundamentals of Data Science"
author: "Student name and id: replace with your name and student id, eg Adam Smith, n12345678 "
date: "October 2026"
output:
  html_document:
    toc: true
    toc_float:
      toc_collapsed: true
    toc_depth: 3
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
options(width = 90)
knitr::opts_chunk$set(echo = TRUE, comment = NA, eval = TRUE)
knitr::opts_chunk$set(fig.width = 4.8, fig.height = 3.6)
options(dplyr.summarise.inform = FALSE, tidyverse.quiet = TRUE, width = 120)
```

# About this assessment

This assessment is about working with and making sense of data using R
and fundamental concepts in probability, statistics, modelling, and visualisation.
It requires you:

- to write and execute code to explore and manipulate data
- to explain things about data, working with data, and data ethics
- to figure things out, including by consulting appropriate sources.

Please check Canvas for more information about this assessment,
including:

- [Instructions](https://canvas.qut.edu.au/courses/23983/pages/assessment-2-instructions "Instructions")
- [When this Assessment is due](https://canvas.qut.edu.au/courses/23983/pages/assessment-2-instructions "Due date")
- [Use of AI and other resources in this Assessment](https://canvas.qut.edu.au/courses/23983/pages/assessment-policy "AI Policy")
- [How will we run and mark your submission](https://canvas.qut.edu.au/courses/23983/pages/assessment-2-marking "Assessment 2: Marking")
- [Errors and corrections to this Assessment](https://canvas.qut.edu.au/courses/23983/pages/assessment-2-errata "Assessment 2: Errata").

------------------------------------------------------------------------

# Part II: Modelling athlete performance (30 marks total)

With Brisbane 2032 on the horizon, many are interested in helping
athletes improve performance. In this part of the assessment, you will
analyse a dataset based on measurements of adolescent tennis players.
The aim of the analysis is to understand the following question:
how can biological measurements be used to predict sporting performance?

To this end, a hypothetical study collected data from 286 participants consisting of 
young tennis players in between the ages of 11-16. Consent for the collection and use 
of their data for this analysis was provided by the participants and their legal guardians.

The data collected include biological data such as:

- `sex`: participant's biological sex
- `height`: participant's height in cm 
- `body_weight`: participant's body weight in kg 
- `fat_mass`: participant's fat mass in kg 
- `lean_mass`: participant's lean mass in kg 

growth related variables such as:

- `years_from_phv`: number of years participant is from their peak height velocity (fastest growing period).
   This is a measure of growth stage, with positive number meaning they are $x$ years past their growth spurt
  and negative numbers meaning they are $x$ years before their growth spurt. 
- `bone_age`: participant's bone age in years. This measures the maturity of their bones. 
- `bone_mineral_density`: participant's bone mineral density in g/cm^3 
- `bone_mineral_content`: participant's in g

The study also tested their upper and lower body performance in terms of their:

- `vertical_jump`: test result for vertical jump task (cm)
- `upper_body_strength`: test result for upper body strength task (cm)
- `hand_grip`: test result for hand grip strength task (kgf)

Finally, a follow-up on this study recorded whether the participants later
went on to win a medal at a regional level tournament (or above) by age 18.
This is recorded in the medal success variable, 

- `medal`: Whether or not the participant went on to win a medal (`"yes"` or `"no"`)

**Note:** The dataset provided is synthetic and was generated based on the 
summary statistics reported by [Almeida-Neto et al. (2021)](https://doi.org/10.1371/journal.pone.0254552).
The `medal` outcome is a teaching scenario rather than an observed outcome from that study.
You are welcome to read the original paper, with its original dataset published 
[here](https://figshare.com/articles/dataset/Raw_Data_xlsx/14755077) under
the [CC BY 40](https://creativecommons.org/licenses/by/4.0/) license.

## Task and Instructions

Your task is to do some initial processing and visualisation of the
data, and then performing regression and classification modelling to see
if the biological measurements are useful in predicting performance and medal
success. 

This is broken down into the sequence of questions listed below, which
largely follows a typical workflow. These questions involve a combination of 
writing code and word responses.
For each question, please write your answer
under the `Answer` heading provided. Code is to be written within the R code 
blocks provided, and word responses are to be written in the plain text blocks
provided. Please do not add additional response blocks for either code or text.

## Clean start

```{r clean-start, class.source="bg-success"}
rm(list = ls())
```

## Load only the specified libraries

```{r load-libraries, warning=FALSE, message=FALSE, class.source="bg-success"}
library(ggplot2)
library(GGally)
library(tidyr)
library(dplyr)
library(readr)
library(caret)
library(rpart.plot)
```

## Load the dataset

```{r load-data, class.source="bg-success"}
sports_data_raw <- read_csv("data/sports_data.csv")
```

------------------------------------------------------------------------

# Question II.1 Processing and Linear Regression Modelling (7 marks)

We begin by exploring whether biological measurement can predict performance 
metrics such as the `vertical_jump`, `upper_body_strength`, or `hand_grip` 
of a tennis player. We first start with some basic exploration of the data.

## Question II.1.1 (1 mark)

Do these tasks in the following order: 


1. List all the column headings of the raw dataset `sports_data_raw` and print out a few rows.

2. Update the `medal` column of the dataset to use a factor type instead of a character type.
  Use the levels `c("no", "yes")`, with `"no"` being the first entry. 
  
3. Store this resulting dataset in a new tibble `sports_data`.


**Note:** We will use the new tibble `sports_data` in the following questions
instead of the raw data `sports_data_raw`. 
The ordering of the factor levels as `c("no", "yes")` is important
in the definition of the logistic regression model in Question 2. This ordering
ensures that the logistic regression model predicts the probability of medal
success (`yes`). 

To make sure you have the correct order, run the following check

```
levels(sports_data$medal)
```
and ensure you get `"no"` before `"yes"`. 

### Answer II.1.1

```{r QII.1.1-code}
# Write your code here 

# Uncomment the following line to check factor levels
# levels(sports_data$medal)
```


## Question II.1.2 (2 marks)

Using the `ggpairs` package, plot pairwise scatter plots of the following columns

- `height`
- `body_weight`
- `lean_mass`
- `fat_mass`
- `years_from_phv`
- `vertical_jump`
- `upper_body_strength`
- `hand_grip`

Then comment on the association between the following pairs of variables:

- `height` and `body_weight` (example showing how body measurements are related)
- `vertical_jump` and `upper_body_strength` (example showing how strength measurements are related)
- `lean_mass` and `hand_grip` (example showing how body and strength measurements are related)

**Note**: A default figure size setting for pairwise scatter plots has been set 
for you, but you are welcome to play with the settings to find a nice visualisation.

### Answer II.1.2

```{r QII.1.2-code, fig.width=10, fig.height=8}
# Write your code here

```

```{asis QII.1.2-text}
Write your response here in plain text

```

## Question II.1.3 (2 marks)

Let's focus on how biological data relates to the upper body strength.
To do so, we will look to fit two linear models using `R`'s linear
regression function `lm`.

Start by fitting a first linear model (Model 1), which predicts `hand_grip` from:

- `sex`
- `years_from_phv`
- `lean_mass`

Then, summarise the fitted model and answer the following questions and
explain your answer in terms of the model summary

- How good is the model fit and how much unexplained variance is there?
- In this model, which variables seem to have a significant impact on
  the measured upper body strengths?

### Answer II.1.3

```{r QII.1.3-code}
# Write your code here
```

```{asis QII.1.3-text}
Write your response here in plain text
```

## Question II.1.4 (2 marks)

Now fit a second linear regression model (Model 2) to predict `hand_grip`  
from all of the other biological measurements + sex. That is,

- `sex`
- `height` 
- `body_weight`
- `fat_mass`
- `lean_mass`
- `years_from_phv`
- `bone_age`
- `bone_mineral_density`
- `bone_mineral_content`

Then summarise the fitted model and answer the following questions and
explain your answer in terms of the model summary

- How good is model fit compared to the first model (Model 1). Was more
  variance explained?
- What happened to the p-value of the `lean_mass` variable compared to Model 1?
  Why do you think this happened?
 
### Answer II.1.4 

```{r QII.1.4-code}
# Write your code here

```


```{asis QII.1.4-text}
Write your response here in plain text

```

# Question II.2 Predicting medal success via logistic regression (10 marks)

Let's now look to predict medal success based on the measured
biological data and performance measures. We will try to fit logistic
regression to classify whether a player will succeed in winning a medal
by age 18 based on the predictors.

## Question II.2.1 (2 marks)

Using `facet_wrap`, make boxplots of each numeric variable. 
Each subplot in the `facet_wrap` should have two boxplots with the x-axis showing
the numeric variable and the y-axis showing whether or not they win a medal. 
Use two different colors for the two different medal categories 
(e.g., boxplot for `medal == yes` is blue and `medal == no` is red).

Then, using these plots, decide if there is indeed some level of separation 
in the measurements between the two medal groups.

### Answer II.2.1

```{r QII.2.1-code}
# Write your code here

```

```{asis QII.2.1-text}
Write your response here in plain text

```

## Question II.2.2 (1 mark)

For each group of `sex = Male` and `sex = Female`, compute the number of 
medal winners and present this as proportion of the total number of males/females.

i.e., present the following numbers for the dataset

- Number of male medal winners
- Number of male medal winners / Total number of male participants 
- Number of female medal winners
- Number of female medal winners / Total number of female participants 

Is there a significant difference between the proportions of male vs female medal winners?

### Answer II.2.2

```{r QII2.2-code}
# Write your code here

```

```{asis QII.2.2-text}
Write your response here in plain text
```

## Question II.2.3 (1 mark)

Using the `createDataPartition` function, partition the full dataset to 
have 80% as training data and 20% as testing. Store these separately. 
For reproducibility, set the random seed to `111` before doing the partitioning
using `set.seed(111)`

### Answer II.2.3

```{r QII.2.3-code}
set.seed(111)
# Write your code here

```

## Question II.2.4 (2 marks)

On the training data only, fit two logistic regression models.

Model 1: predict `medal` from current performance metrics only

`medal` from `vertical_jump`, `upper_body_strength`, and `hand_grip`.

Model 2: predict `medal` from all variables as predictors

And print out their summaries.


### Answer II.2.4

```{r QII.2.4}
# Write your code here 

```

## Question II.2.5 (1 mark)

For each logistic regression model, using a threshold of 0.5, make
predictions for the `medal` success of the test data.
Hint: look into `predict(model, new_data, type = "response").

Display the predicted values of `medal` with each model alongside the
true values for the first 5 rows of the test data

### Answer II.2.5

```{r QII.2.5-code}
# Write your code here
```

## Question II.2.6 (1 mark)

Using a threshold of 0.5, make the confusion matrices for each model.
You can present these as the output of the `table()` function. 

### Answer II.2.6

```{r QII.2.6-code}
# Write your code here

```


## Question II.2.7 (2 marks)

Now, iterating over 40 different values of the threshold between 0 and 1, 
compute the True Positive Rates (TPR) and False Positive Rates (FPR) for each
model over the test dataset. 

Plot these in the form of a receiver operating characteristic (ROC) curve for each model.

What do the ROC curves say about the performance of the two models?

### Answer II.2.7

```{r QII.2.7-code}
# Write your code here

```

```{asis QII.2.7-text}
Write your response here in plain text

```

# Question II.3 Predicting medal success using caret (8 marks)

We will now attempt the classification problem using tree-based models
using the `Caret` package, and compare them to our logistic regression
results.

## Question II.3.1 (1 mark)

First, we need to re-define the factor levels for the `medal` variable. 
Remember that in caret, the convention is reversed compared to the `glm` convention. 
We need the levels to have the success event (`yes`) as the first entry
and the failure (`yes`) event as the second entry.

Relevel the `medal` column to have factors with levels `c("yes", "no')`.
Save your new relevelled tibble as `sports_data_caret` and make sure you use
this dataset for the following questions.

To confirm that you have the result, check the order on your medals column with
```
levels(sports_data_caret$medal)
```

Hint: look into `relevel()`. 


### Answer II.3.1

```{r QII.3.1-code}
# Write your code here

# Uncomment the following to run check
# levels(sports_data_caret$medal)
```


## Question II.3.2 (1 mark)

Again with `createDataPartition` partition the `sports_data_caret` dataset 
into a training and testing dataset. Use 80% of the data as training and 20% of the data as test. 

**Note**: if you're rewriting the partitioning code here, set the random seed
to 111 again for reproducibility using `set.seed(111)`.

### Answer II.3.2 


```{r QII3.2-code}
set.seed(111)

# Write your code here

```

## Question II.3.3 (1 mark)

Fit a logistic regression model in `caret` using all variables as
predictors for `medal`. 

Make sure to only use the training data.

For reproducibility, set the seed to 111 before any training.

### Answer II.3.3

```{r QII.3.3-code}
set.seed(111)

# Write your code here

```

## Question II.3.4 (1 mark)

Train a decision tree using the `rpart` method with all variables as
predictors for `medal`. Make sure to only use the training data.

For the training control, use the following settings

- Do 10-fold cross validation
- Use the ROC metric for cross validation

For reproducibility, set the random seed to 111 before training.

### Answer II.3.4

```{r QII.3.4-code}
set.seed(111)

# write your code here

```

## Question II.3.5 (1 mark)

Using `rpart.plot`, visualise the final decision tree. 
Based on the plot, which variables did the decision tree identify as being important?


### Answer II.3.5

```{r QII.3.5-code}
# Write your code here
```


```{asis QII.3.5-text}
Write your response here in plain text
```

## Question II.3.6 (1 mark)

Now train a random forest using the `rf` method with all variables as
predictors for `medal`. Make sure to only use the training data.

For the training control, use the same settings as the decision tree:

- Do 10-fold cross validation
- Use the ROC metric for cross validation

For reproducibility, set the random seed to 111 before training.

### Answer II.3.6 

```{r QII.3.6-code}
set.seed(111)

# Write your code here

```

## Question II.3.7 (2 marks)

For each of the models from above, i.e., the logistic regression model, 
the decision tree, and the random forest, compute their accuracies and the 
areas under the ROC curve (ROC AUC) using the testing dataset.

Compare and contrast the three models in terms of their performance (accuracy and AUC)
and their interpretability. 

### Answer II.3.7

```{r QII.3.7-code}
# Write your code here

```

```{asis QII.3.7-text}
Write your response here in plain text.

```

# Question II.4 Ethical considerations of performance classification (4 marks)

Suppose that as a sports data scientist, you are tasked with classifying the
performance potentials of young athletes based on their current measurements
(not too dissimilar to what you just did in the previous questions). 

What are the potential ethical issues with this technology? Answer this question 
by identifying two possible stakeholders and two ethically-significant benefits
and harms. 


### Answer II.4

```{asis QII.4-text}
Write your response here in plain text.
```

# Question II.5 (1 mark): Working reproducibly

Follow the reproducible practices covered in the unit (see the Unit Canvas page on this topic and the recommended textbook). 

Your submission must:

- have a separate file for each part 
  (this is part II, so it should be named `Assessment02_Part2.Rmd`)
- render to HTML without error by the Teaching Team;
- use clear code, informative comments, and readable plots;
- avoid hidden dependencies and hard-coded local paths.

------------------------------------------------------------------------

# End of part II
